---
title: "Untitled"
author: "Daniel Abreu"
date: "25/01/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This R program trains a random forest on data from accelerometers. The objective is to train a model to identifying the quality of the activity the participant was doing. After proessing the data, I show that Random Forest model attains good performance in a simulated out-of-sample exercise. The final model is then used to predict data on a entire new (testing) data set.


```{r}
# Load packages
library(caret)
library(randomForest)

#########################################
#Import data
#########################################

setwd("C:/Users/bpu313718/Desktop")
pml_testing <- read.csv("./pml-testing.csv")
pml_training <- read.csv("./pml-training.csv")



#########################################
# Data processing
#########################################

nzv <- nearZeroVar(pml_training) # remove near zero variacnce vars
ptrain <- pml_training[,-nzv] 

mostlyNA <- sapply(ptrain, function(x) mean(is.na(x))) > 0.95 # remove variables that are mostly NA
ptrain <- ptrain[,mostlyNA==F]

ptrain <- ptrain[, -(1:5)] # remove vars that don't make intuitive sense for prediction


# Split training dataset into estimation and validation datasets

set.seed(1)
iTrain <- createDataPartition(y=ptrain$classe, p=0.75, list=F)
ptrain1 <- ptrain[iTrain,]
ptrain2 <- ptrain[-iTrain,]


#########################################
# Model Building
#########################################

# instruct train to use 3-fold CV to select optimal tuning parameters
fitControl <- trainControl(method="cv", number=5, verboseIter=F)

# fit model on ptrain1
fit <- train(classe ~ ., data=ptrain1, method="rf", trControl=fitControl)


#########################################
# Model Evaluation and Selection
#########################################

# predict classe in validation set
preds <- predict(fit, newdata=ptrain2)

# confusion matrix to get the estimate of out-of-sample error
confusionMatrix(as.factor(ptrain2$classe), preds)


#########################################
# Re-training the Selected Model using the full training set
#########################################

fitControl <- trainControl(method="cv", number=5, verboseIter=F)
fit <- train(classe ~ ., data=ptrain, method="rf", trControl=fitControl)


#########################################
# Making Test Set Predictions
#########################################


nzv_t <- nearZeroVar(pml_testing) # remove near zero variacnce vars
ptest <- pml_training[,-nzv_t] 

mostlyNA <- sapply(ptest, function(x) mean(is.na(x))) > 0.95 # remove variables that are mostly NA
ptest <- ptest[,mostlyNA==F]

ptest <- ptest[, -(1:5)] # remove vars that don't make intuitive sense for prediction

# predict on test set
preds <- predict(fit, newdata=ptest)

# confusion matrix to get the estimate of out-of-sample error
confusionMatrix(as.factor(ptest$classe), preds)
```